---
title: "A brief introduction to R"
author: "Arslan A. Zaidi and PingHsun Hsieh"
date: "2024-08-22"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Setting up

In this module, you will receive an introduction to R (<http://www.r-project.org>), an open-source programming language widely used for statistical analysis, and ggplot2, a popular R library for creating plots.

We will use pre-installed R and Rstudio, a user-friendly integrative environment, for this module. In Rstudio, you will see three windows. On your left will be the "Console", where you can run R commands directly. On your top right will be the "Environment", and on your bottom right will be the "Files", which lists all the files and directories inside the project.

Click on "Intro.Rmd", which should open another window on your top left. The .Rmd file is an "Rmarkdown" file, which contains instructions and code needed for this tutorial. The code is shown in blocks encapsulated by three ticks (\`\`\`). You can run the entire block of code by hitting the green "play" arrow at the top right of the code block. Alternatively, you can also run specific lines of code by selecting it and hitting "command/Ctrl + enter". The code will output the results both in the markdown file and in the Console.

In the following section, you will learn basic R code, and students are welcome to partner up, if desired, to help each other.

# Basic R

R can be used as a calculator:

```{r}

# addition; subtraction
# you can write separate commands on the same line by using a semi-colon
2+2; 4-4

# multiplication
8*9

# division
1e3/10

# exponents
2^8

# combination
(2^8) / ((2*3 + 10)/5)

```

You can also assign values to variables and use those.

```{r}

a = 4+5; b = 2
print(a)

a*24

a + b

a^b

# variables can also hold non-numeric
a = "Hello World"; a; print(a)

a = NULL # NULL is a R reserved word representing an undefined value
a = NA  # NA is a R reserved word for missing value

```

**Q1.** Compute $\frac{74+5}{(2.3)^5}$ and assign the name "calculation" to the result. Print "calculation".

```{r}

#code here

```

### Data structures

R has the flexibility of working with different types of data structures. In this section, we will practice the use of vectors, matrices, and data.frames.

```{r}

####### Vectors ############
# vectors are one-dimensional lists of elements
# you can concatenate both characters and numbers using "c()"
a = c(2, "f","hello",10)
a

# you can access specific elements of a vector using index, starting at 0
a[3]; a[2:4]

# changing an element
a[3] = "bye"

#you can also create numeric vectors easily
b = 1:10; d = c(1, 10, 5:6)

```

```{r}

########### Matrices ###########
# A matrix is a two-dimensional object.
# store numbers 1 to 20 in a 5x4 matrix. By default, values are filled by columns.
mat = matrix(c(1:20), nrow = 5, ncol = 4) 
mat

# access specific rows and columns
# 1st and 2nd numbers indicate the indcies for the row and column of interest
mat[2,3]

# access a row
mat[4,]

# access a column
mat[,2]

```

```{r}

########### Data frames ###########
# data frames are more formatted tables with named columns
dat <- data.frame (
  Training = c("Strength", "Stamina", "Other"),
  Pulse = c(100, 150, 120),
  Duration = c(60, 30, 45)
)

dat

# Two ways to access specific columns
dat$Training; dat['Training']

# operate on a column and attach the resulting vector as a new column to the data frame
dat$scaled_Duration = dat$Duration/5

dat
```

### Logical operations

Sometimes we need to perform computations only when certain conditions are met. We can do this using logical operations such as:

-   "==" checks if the values or objects on both sides are equal.
-   "!=" checks if the values or objects on both sides are not equal.
-   "\>" checks if the value/object on the left is greater than that on the right.
-   "\<" checks if the value/object on the left is smaller than that on the right .

```{r}

#create vector of numbers from 1 to 5
a = c(10, 150, 1.5, 30, 65, 300)

#check IF numbers meet the following criterion. 
#this will return a LOGICAL vector
a > 100
a == 10

# Return the indices of the values that meet the criterion.
which(a > 100)

# Return the value(s) of the element(s) that meets the criterion
a[a > 100]

```

We can also specify multiple criterion using the following logical operators:

-   "&" (AND) means both criteria on either side must be met
-   "\|" (OR) means either criteria must be met

```{r}

#AND
a[(a > 100) & (a < 200)]

#OR
a[a < 20 | a > 200]

# Note parantheses are useful for clarity separating operations, but not a must
```

**Q2.** In a fictitious medical study, five male participants were enrolled with height and weight information collected for computing BMI (= mass_kg/height_meter\^2). The goal is to identify individuals with BMI outside the normal range (18.5–24.9). The heights are 180, 163, 200, 174, 171 cm and the weights are 83, 77, 120, 78, 77 kg, respectively. First, create a data frame with two columns, height and weight, and compute a new column BMI. Please use the logical operation to report the indices for individuals with abnormal BMI.

```{r}
# code here

```

### Flow control - loops and conditional statements

In programming, flow control is useful for determining when and how a code block should be executed. In this section, we focus on loops and conditional statements.

A loop is a control structure that repeatedly executes a block of code as long as a specified condition is met. The two most common types of loops are "for" and "while".

```{r}

# "for" loop:
for( i in 1:5 ){      # for values of i from 1 to 5
  print(i^2)          # print the square of i
}

# "while" loop, which keeps running until the condition is NOT met
i = 0                 # set an initial value for the counter
while( i < 5 ){       # exercute the code block while counter is less than 5
  print(i)            # do something with it
  i=i+1               # advance the counter by adding 1; 
                      # if not included, this code block will enter an infinite loop so be very cautious when using a while loop!
}

```

Conditional statements use key words, such as if/else, to check whether certain conditions are met before executing code.

```{r}
a = 10 ; b = 5
if (a < b){
  print ("a is smaller than b")
} else if (a == b){
  print ("a is equal to b")
} else{
    print ("a is greater than b")
  }
}
  
# Combine both loops and conditional statements
for( i in 1:5 ){      # for values of i from 1 to 5
  if (i > 3 ){
     print(i^2)          # print the square of i only when i is greater than 3
  }
}

```

### Functions

A function is a block of code that executes only when it is called. You can pass data, called parameters, into a function, and it can return data as a result. R provides a range of built-in functions and also allows for the creation of user-defined functions.

```{r}

# examples of some built-in functions
max(0, 2, 5); mean(0:10)

# paste() pastes characters or numbers together
paste("This is the square root of 2: ", sqrt(2), sep = "")

# length() returns the length of a vector
length(0:10)

```

```{r}

# my_function: a customized function 
my_function = function(x){
  if(x < 100){
    return(x)
  } else {
    return(NA)
  }
}

my_function(10); my_function(100)

```

One notable built-in function in R is `apply`, which allows you to apply a function to individual rows or columns of a matrix or data frame without explicitly writing a loop.

```{r}

# apply takes three arguments.
# 1st: the matrix or dataframe, 
# 2nd: the axis over which to apply function - 1 for rows, 2 for columns
# 3rd: function you want to apply. it is possible to define your own

# initialize a 5x4 matrix
mat = matrix(c(1:20), nrow = 5, ncol = 4) 

# compute mean over individual rows in mat
apply(mat, 1, mean)

# compute mean over individual rows using a user-defined function
apply(mat, 1, function(x) sum(x)/length(x))   # x is a vector that contains the values of a row.

```

**Q3.** Can you answer Q2 using `apply`without ?

```{r}

```

# Plotting - ggplot2

R has basic plotting functionality. But we will use the `ggplot2` package because it's much more flexible and prettier. You can find more detailed information on [here](https://ggplot2-book.org/introduction.html).

To load a package, use the `library` function. We will use the `mpg` package. It includes information about the fuel economy of popular car models in 1999 and 2008, collected by the US Environmental Protection Agency, <http://fueleconomy.gov>.

```{r}

library(ggplot2)

data(mpg)

head(mpg)

```

Every ggplot2 plot has three key components:

-   Data,

-   A set of aesthetic mappings between variables in the data and visual properties, and

-   At least one layer which describes how to render each observation. Layers are usually created with a geom function.

Let's create a scatterplot:

-   Data: mpg.

-   Aesthetic mapping: engine size mapped to x position, fuel economy to y position.

-   Layer: points.

```{r}

ggplot(mpg, aes(x = displ, y = hwy)) + geom_point()

```

It is important to understand the code structure: data and aesthetic mappings are supplied in `ggplot()`, then layers are added using `+`. This is an important pattern, and as you learn more about ggplot2 you’ll construct increasingly sophisticated plots by adding on more types of components.

Almost every plot maps a variable to x and/or y, so naming these aesthetics is tedious, so the first two unnamed arguments to aes() will be mapped to x and y.

The code specifies `geom_point()` to make a point plot or scatterplot. The `aes()` call can be placed within the \`geom_xxxx\` instead.

In addtion to the default grey background, we can choose from other built-in themes and also edit axis labels and texts.

```{r}

ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  theme_classic() + # change theme
  labs(x = "Engine size", y = "Highway mileage") # axis labels

```

It is easy to customize the color, shape, and transparency of the points as well, either for all of them or grouped one or more factors in the data frame.

```{r}

# To change the color of ALL points, place "color" outside aes()
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy), color = "blue") + 
  theme_classic() + # change theme
  labs(x = "Engine size", y = "Highway mileage") # axis labels

# To change the color of points by some factor, place "color" inside aes()
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy, color = class)) + 
  theme_classic() + # change theme
  labs(x = "Engine size", y = "Highway mileage") # axis labels

```

**Q.** Make a scatterplot between the "cty" variable and "hwy" variable. Label the axes correctly. Color the points by "drv" - the drive type. Describe the relationship between the variables. Is there a drive type that has a higher citywide and highway mileage?

```{r}

#code here

```

You're done with the first part! To proceed, move on to the next part by saving and closing this .Rmd file and opening "PopGen.Rmd". Then go to "Session \> Clear Workspace", followed by "Session \> Restart R". See you there!
